---
title: "Towards a `future` for R"
subtitle: "A fresh approach to parallel computing"
author: "[Nima Hejazi](http://nimahejazi.org)"
date: "`r lubridate::now()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "custom.css"]
    nature:
      highlightStyle: zenburn
      highlightLines: true
      countIncrementalSlides: false
---

```{r knitr_setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 4.5, dpi = 300,
                      fig.cap = "", fig.align = "center")
showtext::showtext.opts(dpi = 300)
```

# Accessing these slides

--

### View online:

* shortened URL: [https://goo.gl/T47hs6](https://goo.gl/T47hs6)

--

### Build locally:

_You'll need to clone the `git` repo of source materials and then build the
slide deck (HTML)._

```{bash, eval=FALSE}
git clone https://github.com/nhejazi/talk_futuRe-intro.git ~/futuRe-intro
cd ~/futuRe-intro
make  # requires the R package "xaringan"
open ./2017_futuRe_intro.html
```

???

- This talk will focus on introducing an ecosystem of R packages centered around
futures, which collectively provide tools for __asynchronous__ computation in R.

---
class: inverse, center, middle

# R's `future` is now: Core concepts

---

# A `promise` of a better `future`

## So, what are `future`s?

--

"A `future` is an abstraction for a value that may be available at some point in
the future. The state of a `future` can either be unresolved or resolved. As
soon as it is resolved, the value is available instantaneously. If the value is
queried while the `future` is still unresolved, the current process is blocked
until the `future` is resolved." - excerpted from "A Future for R: A
Comprehensive Overview"

--

- `future` - an object corresponding to an expression to be evaluated at a later
  point in time, when its result is needed.

--

- `promise` - a _lazy evaluation mechanism_ that results in a given expression
  being evaluated only if its value is requested.

--

```{r future-explicit, eval=FALSE}
f <- future({ expr })  # creates a future
v <- value(f)  # gets the value of the future (blocks if not yet resolved)
```

--

```{r future-implicit, eval=FALSE}
v %<-% { expr }  # creates a future and a promise to its value
```

???

- Both futures and promises are general programming concepts, not specific to R.
- Futures allow for the computation of procedures/expressions to be postpoined
  to some later time.
- Promises are a mechanism that ensures that futures will be evaluated only when
  they are needed. These work by a mechanism of delayed computation. See the R's
  `delayedAssign` base function or Python's `dask` library.

---

# How to `plan` your `future`

There are many strategies for evaluating a `future`. Within the nomenclature of
the `future` ecosystem, these are called `plan`s.

--

### Synchronous (non-parallel) evaluation

- `sequential`: sequentially and in the current R process

- `transparent`: as sequential but with early signaling (for debugging)

--

### Asynchronous (parallel) evaluation

- `multicore`: forked R processes on current machine (_n.b._, not supported on
Windows)

- `multisession`: background R sessions on current machine

- `multiprocess`: `multicore` if supported; otherwise, `multisession`

- `cluster`: external R sessions on current, local, and/or remote machines

- `remote`: simple access to remote R sessions

???

- ...

---

# The `future` ecosystem

```{r install_pkgs, message=FALSE, eval=FALSE}
install.packages(c("future", "doFuture", "future.batchtools"))
source("http://callr.org/install#HenrikBengtsson/BiocParallel.FutureParam")
```

--

- `doFuture`: "A Universal Foreach Parallel Adaptor using the Future API"
  - `%dopar%` adaptor allows the `foreach` package to work with any type of
    `future`.
  - Allows `future`s to work with HPC clusters and the `BiocParallel` package.
  - Plug-in replacement for packages like `doParallel`, `doSNOW`, `doMPI`, etc.
  - Check it out: https://github.com/HenrikBengtsson/doFuture

--

- `future.batchtools`: "A Future API for Parallel and Distributed Processing"
  - Leverage HPC clusters just by a switching settings - your code stays exactly
    as it is!
  - Supports several HPC job schedulers: TORQUE, Slurm, SGE, LSF, OpenLava
  - Check it out: https://github.com/HenrikBengtsson/future.batchtools

--

- `BiocParallel.FutureParam`: "A `BiocParallelParam` Class for Futures"
  - `FutureParam` allows `BiocParallel` to work with any type of `future`.
  - Allows `future`s to operate seamlessly within the
    [Bioconductor](http://www.bioconductor.org/) ecosystem.
  - Plug-in replacement for `MulticoreParam()`, `BatchJobsParam`, etc.
  - Check it out: https://github.com/HenrikBengtsson/BiocParallel.FutureParam

???

- `doFuture` resolves issues with global variables that normally arise with
  `foreach` (recall the `.export` requirement).
- An abundance of documentation is available for these extension packages - best
  to consult these for more information.

---
class: inverse, center, middle

# `future`s in Action

## There ain't nothin' like a test drive...

---

# Computing with `future`s I

```{r prelims_pkgs, message=FALSE, echo=FALSE}
library(future)
library(listenv)
#future::availableCores()
```

Let's try computing via a nested for-loop with `future`s.

--

```{r nested-for-future}
future_varfuns <- function(x, n = 10, times = 10, fun = "mean") {
  # over the columns of a 2-d array, compute the mean of FUN within a sample
  var_meanfuns <- listenv::listenv()
  for (j in seq_len(ncol(x))) {  # loop over variables/columns
    var_meanfuns[[j]] %<-% {
      samp_fun <- listenv::listenv()
      for (i in seq_len(times)) {
        samp_fun[[i]] %<-% match.fun(fun)(x[sample(nrow(x), n), j])
      }
      as.list(mean(unlist(samp_fun)))
    }
  }
  return(var_meanfuns)
}
```

--

```{r compute-nested-for, message=FALSE, warning=FALSE}
plan(list(multiprocess, sequential))
result <- future_varfuns(x = replicate(4, rnorm(100)))
unlist(as.list(result))
```

???

- Could have reversed which loop gets parallelized by flipping `multiprocess`
  and `sequential`.
- If submitting to HPC, could have invoked the scheduler (for nodes), then core
  parallelization (e.g., `plan(list(batchtools_slurm, multiprocess))`)

---

# Computing with `future`s II

The `origami` package provides a highly flexible way to perform cross-validation
of (nearly) _any_ R function, relying on `future`s for (a)synchronous
computation.

```{r load_origami, message=FALSE}
library(origami)
data(mtcars)
folds <- make_folds(mtcars)
```

```{r setup-origami, echo=FALSE, message=FALSE}
cv_lm <- function(fold, data, reg_form) {
  # get name and index of outcome variable from regression formula
  out_var <- as.character(unlist(str_split(reg_form, " "))[1])
  out_var_ind <- as.numeric(which(colnames(data) == out_var))

  # split up data into training and validation sets
  train_data <- training(data)
  valid_data <- validation(data)

  # fit linear model on training set and predict on validation set
  mod <- lm(as.formula(reg_form), data = train_data)
  preds <- predict(mod, newdata = valid_data)

  # capture results to be returned as output
  out <- list(coef = data.frame(t(coef(mod))),
              SE = ((preds - valid_data[, out_var_ind])^2))
  return(out)
}
```

--

```{r future-sequential, message=FALSE, warning=FALSE}
plan(sequential)
time_seq <- system.time({
  results_seq <- cross_validate(cv_fun = cv_lm, folds = folds,
                                data = mtcars, reg_form = "mpg ~ .")
})
```

--

```{r future-parallel, message=FALSE, warning=FALSE}
plan(multisession)
time_ms <- system.time({
  results_ms <- cross_validate(cv_fun = cv_lm, folds = folds,
                               data = mtcars, reg_form = "mpg ~ .")
})
```

???

- ...

- ...

---

# Tooling with `future`s

### R/`origami`

- Core function is a wrapper around `future_lapply`, allowing for a variety of
  cross-validation schemes to be applied with delayed computations.
- Check it out: https://github.com/jeremyrcoyle/origami
- status: _released_ - just grab it from CRAN!

--

### R/`cvma`

- Nested cross-validation for summarizing associations in multivariate outcome
  settings (parallelized via `future_lapply`).
- Check it out: https://github.com/benkeser/cvma
- status: _in early development_ - no guarantees yet.

--

### R/`methyvim`

- Nonparametric analysis of DNA differential methylation (combines `future`,
  `doFuture`, `BiocParallel.FutureParam`, `BiocParallel`).
- Check it out: https://github.com/nhejazi/methyvim
- status: _in development_ - on Bioconductor (v3.6), to be released 31 October.

???

- For a look at how `future`, `doFuture`, and `Biocparallel.Futureparam` can be
combined, `methyvim::set_parallel`.
- ...


---
class: inverse, center, middle

# Thanks!

twitter/[@nshejazi](https://twitter.com/nshejazi),
github/[@nhejazi](https://github.com/nhejazi)

[nimahejazi.org](http://nimahejazi.org/), [stat.berkeley.edu/~nhejazi](https://www.stat.berkeley.edu/~nhejazi/)

Slides created via the R package
[**xaringan**](https://github.com/yihui/xaringan).

Powered by [remark.js](https://remarkjs.com),
[**knitr**](http://yihui.name/knitr), and
[R Markdown](https://rmarkdown.rstudio.com).
